Language and vision in conceptual processing: Multilevel analysis and statistical power

### 2022-10-24 manuscript/manuscript.tex
24102022

\abstract{%
Research has suggested that conceptual processing depends on both language-based and vision-based information. We tested this interplay at three levels of the experimental structure: individuals, words and tasks. To this end, we drew on three existing, large data sets that implemented the paradigms of semantic priming, semantic decision and lexical decision. We extended these data sets with measures of language-based and vision-based information, and analysed how the latter variables interacted with participants' vocabulary size and gender, and also with presentation speed in the semantic priming study. We performed the analysis using mixed-effects models that included a comprehensive array of fixed effects---including covariates---and random effects. First, we found that language-based information was found to be more important than vision-based information. Second, in the semantic priming study---whose task required distinguishing between words and nonwords---, both language-based and vision-based information were more influential when words were presented faster. Third, a `task-relevance advantage' was identified in higher-vocabulary participants. Specifically, in lexical decision, higher-vocabulary participants were more sensitive to language-based information than lower-vocabulary participants. In contrast, in semantic decision, higher-vocabulary participants were more sensitive to word concreteness. Fourth, we demonstrated the influence of the analytical method on the results. These findings support the interplay between language and vision in conceptual processing, and demonstrate the influence of measurement instruments on the results. Last, we estimated the sample size required to reliably investigate various effects. We found that 300 participants were sufficient to examine the effect of language-based information contained in words, whereas more than 1,000 participants were necessary for the effect of vision-based information and for the interactions of both former variables with vocabulary size, gender and presentation speed. In conclusion, this power analysis reveals the need to increase sample sizes for research on perceptual simulation and individual differences.
}

@@

We entertained two hypothesis regarding the interaction between language-based information and participants' gender: (a) that the language system would be more important in female participants than in males (Burman et al., 2008; Hutchinson \& Louwerse, 2013; Jung et al., 2019; Ullman et al., 2008), and (b) that this interaction effect would be absent, as a recent review suggested that gender differences are negligible in the general population (Wallentin, 2020).

@@

To our knowledge, no previous studies have investigated the interaction between vision-based information and participants' gender. We entertained two hypothesis. Our first hypothesis was that this interaction would stand opposite to the interaction between language and gender. That is, if female participants were to present a greater role of language-based information, male participants would present a greater role of vision-based information, thereby compensating for the disadvantage on the language side. Our second hypothesis was the absence of this interaction effect (see Wallentin, 2020).

>>>>

\abstract{%
Research has suggested that conceptual processing depends on both language-based and vision-based information. We tested this interplay at three levels of the experimental structure: individuals, words and tasks. To this end, we drew on three existing, large data sets that implemented the paradigms of semantic priming, semantic decision and lexical decision. We extended these data sets with measures of language-based and vision-based information, and analysed how the latter variables interacted with participants' vocabulary size and gender, and also with presentation speed in the semantic priming study. We performed the analysis using mixed-effects models that included a comprehensive array of fixed effects---including covariates---and random effects. First, we found that language-based information was more important than vision-based information. Second, in the semantic priming study---whose task required distinguishing between words and nonwords---, both language-based and vision-based information were more influential when words were presented faster. Third, a `task-relevance advantage' was identified in higher-vocabulary participants. Specifically, in lexical decision, higher-vocabulary participants were more sensitive to language-based information than lower-vocabulary participants. In contrast, in semantic decision, higher-vocabulary participants were more sensitive to word concreteness. Fourth, we demonstrated the influence of the analytical method on the results. These findings support the interplay between language and vision in conceptual processing, and demonstrate the influence of measurement instruments on the results. Last, we estimated the sample size required to reliably investigate various effects. We found that 300 participants were sufficient to examine the effect of language-based information contained in words, whereas more than 1,000 participants were necessary to examine the effect of vision-based information and the interactions of both former variables with vocabulary size, gender and presentation speed. In conclusion, this power analysis reveals the need to increase sample sizes when conducting research on perceptual simulation and individual differences.
}

@@

We entertained two hypotheses regarding the interaction between language-based information and participants' gender: (a) that the language system would be more important in female participants than in males (Burman et al., 2008; Hutchinson \& Louwerse, 2013; Jung et al., 2019; Ullman et al., 2008), and (b) that this interaction effect would be absent, as a recent review suggested that gender differences are negligible in the general population (Wallentin, 2020).

@@

To our knowledge, no previous studies have investigated the interaction between vision-based information and participants' gender. We entertained two hypotheses. Our first hypothesis was that this interaction would stand opposite to the interaction between language and gender. That is, if female participants were to present a greater role of language-based information, male participants would present a greater role of vision-based information, thereby compensating for the disadvantage on the language side. Our second hypothesis was the absence of this interaction effect (see Wallentin, 2020).
